Chaotic mixing of tidal debris

### 2015-06-10 ms.tex
adding text

% packages
\usepackage{amssymb,amsmath,amsbsy}
\usepackage{booktabs}
\usepackage{bbold}
\usepackage{mathrsfs}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage[yyyymmdd]{datetime}
\renewcommand{\dateseparator}{-}

@@

In order to explore the question of how chaos manifests over timescales much shorter than that predicted from generic chaos indicators, we must first understand the structure and behavior of orbits in gravitational potentials.

@@

An orbit in an $N$ degree of freedom (dof) Hamiltonian, $H$, is a set of $2N$ quasi-periodic time series, 
\begin{equation}
(w_1(t),...,w_{2N}(t)) = (q_1(t),...,q_{N}(t),p_1(t),...,p_{N}(t)) \label{eq:coords}
\end{equation}
where $q_i$ and $p_i$ are conjugate coordinates in the sense that
\begin{align}
	\dot{p}_i &= -\frac{\partial H}{\partial q_i}\\
	\dot{q}_i &= \frac{\partial H}{\partial p_i}
\end{align}
for all $t$. If bounded, the motion in any component, $w_i(t)$, can be represented as a Fourier sum,
\begin{equation}
	w_i(t) = \sum_k^\infty a_k \, e^{i\,\omega_k\,t} \label{eq:fourier}
\end{equation}
where the $a_k$ are complex amplitudes.

A \emph{regular orbit} is a set of such time series that can be transformed to a special set of canonical coordinates known as angle-action coordinates. In these coordinates, the position variables are angles, $\boldsymbol{\theta}$, that increase linearly with time with rates set by $N$ constant, fundamental frequencies, $\boldsymbol{\Omega} = (\Omega_1, ..., \Omega_N)$. The frequency of a Fourier component in the sum for any individual component of motion (the $\omega_k$ in Equation~\ref{eq:fourier}) are just linear, integer combinations of the fundamental frequencies, $\boldsymbol{\Omega}$ --- that is, for a regular orbit, any Fourier component frequency may be written
\begin{align}
	\omega_k &= \boldsymbol{n_k} \cdot \boldsymbol{\Omega} \label{eq:fourierfreq}
\end{align} %; \, (\boldsymbol{n}_k \in \mathbb{Z}^3)
where $\boldsymbol{n}_k$ is a vector of $N$ integers. The conjugate momentum coordinates --- the actions,  $\boldsymbol{J}$ --- are constants of motion. Even stronger, the actions are isolating integrals and any pair are in involution such that
\begin{equation}
	[J_i, J_j] = 0
\end{equation}
where $[\cdot,\cdot]$ is the Poisson bracket. This implies that for an $N$ dof system, a regular orbit has $N$ independent constants of motion and the motion is therefore restricted to an $N$-dimensional manifold embedded in the 2$N$ dimensional phase space. In fact, the topology of angle-action space is toroidal and any regular orbit in an $N$ dof Hamiltonian can be understood as motion on the surface of an $N$-torus. Each set of actions, $(J_1,...,J_N)$ (or frequencies), uniquely labels a torus, and regular orbits are sometimes referred to in terms of their orbital tori.

@@

The frequencies of a generic, non-resonant orbit are incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} Conditionally-periodic orbits uniformly cover the surface of their orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ then the orbit is a resonant orbit. For resonant orbits, the existence of an additional \emph{local} constant of motion reduces the dimensionality of the orbit. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic 2D surfaces.

\subsection{Orbits in non-integrable potentials}

\todo{APW}{take cartoons from talk and include here for description}

The orbit structure of non-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a complete description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99} and the stable regions are surrounded by stochastic layers where chaotic motion occurs. Chaotic orbits are characterized by exhibiting random behavior despite being generated by entirely deterministic equations of motion. Stochastic layers form gaps in action space where tori are destroyed; the orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies are not constant and diffuse chaotically with time. For small values of $\epsilon$, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}.

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori caused by the onset of chaotic motion. [... how and where to end this section?]

@@

[??]

@@

Small ensembles of (conditionally-periodic) regular orbits will reach a fully phase-mixed state in a timescale  $\approx\sigma_\Omega^{-1}$, where $\sigma_\Omega$ is the dispersion in frequency-space of the ensemble --- each individual orbit will fill the surface of its torus ergodically, but the ensemble will only mix due to sheering from slight variations in their fundamental frequencies \citep{merritt96, helmi99}. Ensembles of chaotic orbits behave quite differently. Over long times, a chaotic orbit in an autonomous Hamiltonian will ergodically mix over the energy hypersurface (a 5D surface, compared to the 3D stable tori). Generically, a small ensemble of chaotic orbits will lose coherence much faster than for regular orbits, however analytically computing the rate and amount of spreading is very difficult even for simple potentials \cite[see, e.g.,][]{chirikov??}. Certain ensembles may be trapped between resonances for long times and thus remain in a state of pseudo-coherence for long times, whereas others may rapidly diffuse \citep[see, e.g.,][]{merritt96}. This phenomenon is generic to potentials that are not globally stochastic in wich regions of stability and chaos are intermingled throughout phase-space.

@@

Since all known tidal streams have been discovered as kinematic over-densities of stars, this implies (1) [something about accretion history], (2) the long, thin tidal streams we see may represent a sampling from ``special'' orbits in the Galactic potential where the mixing time is long, and (3) previous assumptions that connect the length and size of streams to their evolution time may be invalid.

@@

\subsection{Potential}

Why not disk + bulge? What do we expect when we add?

@@

\section{Conclusions}\label{sec:conclusions}
We have shown that ...

@@

% Figure 1
% TODO: should I show density? cool 3D render of density?
\begin{figure*}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/potential.pdf}
\caption{Equipotential contours for the triaxial NFW potential considered in this work. There are eight contour levels evenly spaced and linear in the value of the potential. } \label{fig:potential}
\end{center}
\end{figure*}

>>>>

% packages
\usepackage{amssymb,amsmath,amsbsy}
\usepackage{booktabs}
\usepackage{bbold}
\usepackage{mathrsfs}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage[yyyymmdd]{datetime}
\renewcommand{\dateseparator}{-}
\usepackage{lscape}
\usepackage{rotating}

@@

In order to explore the question of how chaos manifests over timescales much shorter than that predicted from generic chaos indicators, we must first understand the structure and behavior of orbits in complex gravitational potentials. An orbit in an $N$ degree of freedom (dof) Hamiltonian, $H$, is a set of $2N$ quasi-periodic time series, 
\begin{equation}
(w_1(t),...,w_{2N}(t)) = (q_1(t),...,q_{N}(t),p_1(t),...,p_{N}(t)) \label{eq:coords}
\end{equation}
where $q_i$ and $p_i$ are conjugate coordinates in the sense that
\begin{align}
	\dot{p}_i &= -\frac{\partial H}{\partial q_i}\\
	\dot{q}_i &= \frac{\partial H}{\partial p_i}
\end{align}
for all $t$. If bounded, the motion in any component, $w_i(t)$, can be represented as a Fourier sum,
\begin{equation}
	w_i(t) = \sum_k^\infty a_k \, e^{i\,\omega_k\,t} \label{eq:fourier}
\end{equation}
where the $a_k$ are complex amplitudes.

A \emph{regular orbit} is a set of such time series that can be transformed to a special set of canonical coordinates known as angle-action coordinates. In these coordinates, the position variables are angles, $\boldsymbol{\theta}$, that increase linearly with time with rates set by $N$ constant, fundamental frequencies, $\boldsymbol{\Omega} = (\Omega_1, ..., \Omega_N)$. The frequency of a Fourier component in the sum for any individual component of motion (the $\omega_k$ in Equation~\ref{eq:fourier}) are just linear, integer combinations of the fundamental frequencies, $\boldsymbol{\Omega}$ --- that is, for a regular orbit, any Fourier component frequency may be written
\begin{align}
	\omega_k &= \boldsymbol{n_k} \cdot \boldsymbol{\Omega} \label{eq:fourierfreq}
\end{align} %; \, (\boldsymbol{n}_k \in \mathbb{Z}^3)
where $\boldsymbol{n}_k$ is a vector of $N$ integers. The conjugate momentum coordinates --- the actions,  $\boldsymbol{J}$ --- are constants of motion. Even stronger, the actions are isolating integrals and any pair are in involution such that
\begin{equation}
	[J_i, J_j] = 0
\end{equation}
where $[\cdot,\cdot]$ is the Poisson bracket. This implies that for an $N$ dof system, a regular orbit has $N$ independent constants of motion and the motion is therefore restricted to an $N$-dimensional manifold embedded in the 2$N$ dimensional phase space. The topology of angle-action space is toroidal and any regular orbit in an $N$ dof Hamiltonian can be understood as motion on the surface of an $N$-torus. Each set of actions, $(J_1,...,J_N)$ (or frequencies), uniquely labels a torus, and regular orbits are sometimes referred to in terms of their orbital tori.

@@

@@

The frequencies of a generic, non-resonant orbit are incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is a resonant orbit. For resonant orbits, the existence of an additional \emph{local} constant of motion reduces the dimensionality of the orbit. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic 2D surfaces. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all non-resonant orbits are regular.

\subsection{Orbits in near- and non-integrable potentials}

The orbit structure of near-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a complete description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99}, and these layers regions are then generically surrounded by stochastic layers where chaotic motion occurs. Chaotic orbits are characterized by exhibiting random behavior despite being generated by entirely deterministic equations of motion. Stochastic layers form gaps in action space where tori are destroyed; the orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies are not constant and diffuse chaotically with time.

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel). Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that will librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined in some way, it may behave nearly regular for long periods of time.

For small values of $\epsilon$, many regular orbits survive and only small chaotic regions are introduced, especially where resonances overlap. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Figure~\ref{fig:cartoons} (right panel) qualitatively shows this phenomenon --- the resonant and resonant-layer orbits may still be regular, but many or all of the conditionally-periodic orbits can be chaotic. As the perturbation strength increases, next the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \cite[for a more quantitative illustration of this transition from integrability to global chaos, see Figure~9 in][]{valluri98}.

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori caused by the onset of chaotic motion.

@@

Chaotic orbits ...

Small ensembles of (conditionally-periodic) regular orbits will reach a fully phase-mixed state in a timescale  $\approx\sigma_\Omega^{-1}$, where $\sigma_\Omega$ is the dispersion in frequency-space of the ensemble --- each individual orbit will fill the surface of its torus ergodically, but the ensemble will only mix due to sheering from slight variations in their fundamental frequencies \citep{merritt96, helmi99}. Ensembles of chaotic orbits behave quite differently. Over long times, a chaotic orbit in an autonomous Hamiltonian will ergodically mix over the energy hypersurface (a 5D surface, compared to the 3D stable tori). Generically, a small ensemble of chaotic orbits will lose coherence much faster than for regular orbits, however analytically computing the rate and amount of spreading is very difficult even for simple potentials \cite[see, e.g.,][]{chirikov??}. Certain ensembles may be trapped between resonances for long times and thus remain in a state of pseudo-coherence for long times, whereas others may rapidly diffuse \citep[see, e.g.,][]{merritt96}. This phenomenon is generic to potentials that are not globally stochastic in which regions of stability and chaos are intermingled throughout phase-space.

\todo{apw}{concluding paragraph...}

@@

@@

\subsection{Existence of streams}

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have found that

\subsection{Cautions}

Since all known tidal streams have been discovered as spatial over-densities of stars, this implies (1) [something about accretion history], (2) the long, thin tidal streams we see may represent a sampling from ``special'' orbits in the Galactic potential where the mixing time is long, and (3) previous assumptions that connect the length and size of streams to their evolution time may be invalid.

@@

[discussion of potential choice?]
Why not disk + bulge? What do we expect when we add?

@@

\section{Summary and Conclusions}\label{sec:conclusions}
We have shown that ...

@@

% Figure ??
\begin{sidewaysfigure}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/cartoons.pdf}
\caption{} 
\label{fig:cartoons}
\end{center}
\end{sidewaysfigure}

% Figure ??
\begin{figure*}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/potential.pdf}
\caption{Equipotential contours for the triaxial NFW potential considered in this work. There are eight contour levels evenly spaced and linear in the value of the potential. \todo{apw}{should I show density as well or instead?}} \label{fig:potential}
\end{center}
\end{figure*}
